Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-29 paper/paper.tex
latest - 29th

\begin{abstract} 
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs. Our approach is to analyze
programs  for liveness  of  heap-allocated data.   The  result of  the
analysis is  used to preserve  only live data---a subset  of reachable
data---during garbage  collection.  The result  is an increase  in the
garbage reclaimed and  a  reduction in  the peak memory requirement 
of programs.   While this  technique has already  been shown  to yield
benefits for  eager first-order  languages, the  lack of  a statically
determinable execution  order and  the presence  of closures  pose new
challenges  for lazy  languages.  These  requires changes  both in  the
liveness analysis itself and in the design of the garbage collector.

@@

While  the runtime  system  of most  functional  languages includes  a
garbage  collector  (GC)  to  efficiently  reclaim  memory,  empirical
studies on Scheme~\cite{karkare06effectiveness} and, more importantly,
on  Haskell~\cite{rojemo96lag}  programs  have shown  that  GCs  leave
uncollected a  large number of  memory objects that are  reachable but
not live (here {\em live} means  the object can potentially be used by
the program  at a  later stage).  This  results in  unnecessary memory
retention.

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis has  two  parts.   First, we  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function  calls during  analysis.  Second,  the concrete  demand on  a
function  body is  obtained through  a conservative approximation
similar to 0-CFA~\cite{Shivers:1988} that combines the  demands on all the calls
to  the function.   The result  of the  analysis is  an annotation  of
certain program points with  deterministic finite-state automata (DFA)
capturing the liveness of variables at these points.  Depending on the
program point where garbage collection has been triggered, the garbage
collector consults a  set of automata to  restrict reachability during
marking.  This  results in  an increase in  the garbage  reclaimed and
consequently in fewer garbage collections.

While this idea has been shown  to be effective for a first-order {\em
  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
the technique is not possible for lazy languages, where heap-allocated
objects may  include closures (runtime representations  of unevaluated
expressions).   Firstly, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  Secondly,  for liveness-based
garbage collection to  be effective, we need  to extend liveness-based
collection to closures apart from evaluated data.  Since a closure can
escape the scope in which it was created, it has to carry the liveness
information of  its free variables. Moreover,  as execution progresses
and  possible  future uses  are  eliminated,  we update  the  liveness
information  in a  closure with  a more  precise version.   Because of
these reasons,  the garbage collector also  becomes significantly more
complicated than a liveness-based collector for a eager language.

\warning{Experiments  with  a   single  generation  copying  collector
  (Section~\ref{sec:experiments})  confirm  the  expected  performance
  benefits.   Liveness-based  collection  results in  an  increase  in
  garbage reclaimed.   As a  consequence, there is  a decrease  in the
  number of  collections (up to 45\%)  and a decrease (up  to 88\%) in
  the number  of {\em dragged}  cells (cells  that survive a  GC after
  their  last use).  Besides,  there  is a  reduction  in the  overall
  execution time for some programs.  Our experiments show that a mixed
  strategy, where we  do liveness based GC up to  the closure boundary
  and reachability based GC for  the arguments (free variables) of the
  closure,  gives us  similar benefits  as a  fully liveness  based GC
  (where  the  arguments  of  the closure  are  also  collected  using
  liveness) with lesser overhead (Section~\ref{sec:strategies}).}

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\input{fig_motiv}
Figure~\ref{fig:mot-example} shows an example program and the state of
the heap at  the program point $\pi$, i.e. just  before the evaluation
of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
node either  represents atomic  values ($\NIL$,  integers etc.),  or a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent  variables  or fields.  The figure shows
the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
triggered by the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.

@@

%\comment{  
Consider  the program in  Figure~\ref{fig:mot-example2}.  As
  mentioned earlier,  a lazy  evaluation of  the $\LET$  expression at
  $\pi_1$ creates a closure for $(length~x)$ instead of evaluating it.
  Since the  closure may escape the  scope in which it  is created, it
  carries a copy  of $x$ within itself.   We treat the copy  of $x$ in
  the closure as being separate from the $x$ introduced by the $\LET$,
  and call  it a \emph{closure variable}.   For liveness calculations,
  such variables are distinguished from variables introduced by \LET s
  and  function arguments  (called \emph{stack  variables} since  they
  reside in the activation stack). Notationally, we distinguish
  a closure variable from its corresponding stack variable by
  subscripting it with the program point where the closure was created\footnote{Mutiple occurrences of the same variable in an
    application are further distinguished by their positions in the
    application.}.

@@

  Closure  variables get  their  liveness values  through  a chain  of
  dependences beginning at  a variable at an evaluation  point.  As an
  example,  the  variable $z$  returned  at  $\psi_2$ depends  on  $y$
  through  the  expression $(+~y~1)$.   $y$  in  turn depends  on  the
  closure  variable  $x_{\pi_1}$   through  $(length~x_{\pi_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!z \leftarrow
  (+~y~1),\,  y  \leftarrow   (length~x_{\pi_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $z$  is  evaluated   at  $\psi_2$  due  to  the
  expression $\SRETURN~z$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $z$ which  will impart  a liveness  to
  $x_{\pi_1}$.  Other dependence chains which result in a liveness for
  $x_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!y       \leftarrow
  (length~x_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!w \leftarrow (/~u~y),\,\,
  y  \leftarrow  (length~x_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $x_{\pi_1}$ to be
  a union of the liveness arising  out of these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $y$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $x_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $z$ at $\psi_2$, it can safely consider only the liveness
  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!z \leftarrow
  (y+1),\,   y  \leftarrow   (length~x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

%% \begin{figure*}[t!]
%% \begin{center}\footnotesize
%% \renewcommand{\arraystretch}{1.5}

%% \begin{tabular}{|c|c|c|}
%% \hline
%% Premise & Transition & Rule name \\ 
%% \hline

%% \hline
%% \makecell[t]{$GC(\rho_1, S, \heap_1, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
%% \sc{car-clo-bang}
%% \\
%% \hline
%% \makecell[t]{$GC(\rho_1, S, \heap_1, \Lv, \sigma) = (\rho, S, \heap)$,}\\ $\heap(\rho(x))
%% \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
%% \sc{car-clo}\\
%% \hline
%% \end{tabular}

%% \medskip

%% \begin{tabular}{|c|c|c|}
%% \hline
%% Premise & Transition & Rule name \\ 
%% \hline

%% \hline

%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
%%   \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(x) :=
%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
%% \\
%% \hline
%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
%%   \rho', S, \heap[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } (\ell_1, \ell_2)$ & \makecell{$\rho, (\rho', z, e,
%%   \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
%%   \rightsquigarrow \rho', S, \heap[\rho'(z) := \heap(\ell_1)], e, \cred{\sigma'}$}      &
%% \sc{car-whnf} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
%% \sc{car-clo}
%% \\

%% \hline
%% $\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
%% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
%% \sc{car-1-clo} \\

%% \hline
%% $\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap,
%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, \heap[\rho'(z)
%%     := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
%% \sc{prim-whnf} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
%%   \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-1-clo} \\
%% \hline
%% $\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
%%   S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-2-clo} \\
%% \hline
%% {$\mathit{g}~\mbox{defined as}$
%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, \heap,
%%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
%%     \rho(\myvec{x})], S, \heap, e_{\mathit{g}}, \cred{\sigma}$}      &
%% \sc{funcall} \\
%% \hline
%% $\ell$ is a new location& \makecell{$\rho, S, \heap, (\LET~x\leftarrow
%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
%%   \ell]\rangle], e, \cred{\sigma}$} &
%% \sc{let} \\ 
%% \hline
%% $\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
%% \hline
%% $\heap(\rho(x)) = 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$   $\rightsquigarrow
%% \rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
%% \hline
%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
%% \cred{\clazy\sigma}$}
%% &
%% \sc{if-clo} \\
%% \hline
%% {$\heap(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
%%   (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(z) := v], e,
%%   \cred{\sigma'}$} &
%% \sc{return-whnf}\\
%% \hline
%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
%%   (\psi:\SRETURN~x), \cred{\sigma}$  $
%%   \rightsquigarrow$
%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
%%   \cred{\sigma}$} &
%% \sc{return-clo} \\
%% \hline
%% \end{tabular}
%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
%% \end{center}
%% \vskip -5mm
%% \end{figure*}

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.  %% \warning{While  the proof  is on
  %% the broad lines of  \cite{asati14lgc}, there are several differences
  %% due to the  lazy semantics of the language in  this paper.}
Here are
the ideas behind the proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
  collection  just  before  each  $\rightsquigarrow$  transition.  The
  simulated garbage collection starts from  the root-set and inserts a
  special value  $\bot$ at each location  in the heap that  contains a
  reference that is reachable but not live. Any attempt to dereference
  such  locations during  the  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for \sc{car-clo-bang} and
  \sc{car-clo-bang} \label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item   We   extend    the   abstract   machine   state
  $\rho,S,\heap,e$   to  $\rho,S,\heap,e,\sigma$. 
We  call such a state  a \emph{minefield state}.
Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
  $e$, i.e., it arises
  from the  actual sequence of function  calls that led
  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
  %% during    liveness     analysis    is     a    static
  %% over-approximation   of   $\sigma$   that   considers
  %% \emph{all  possible calls}  that  could  lead to  the
  %% evaluation of  $e$.
  The  information in  a suspended
  evaluation  context  on  the   stack  $S$  is  also
  similarly augmented with its  demand.  Thus a stacked
  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
  %% modification  of  the   small  step  semantics  which
  %% carries the extra information  is shown in the bottom
  %% table of Figure~\ref{fig:minefield-semantics}.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This represents the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item \label{gc-1}Using the current  evaluation context $\rho, \stk,
    \heap,   e,  \sigma$   and   the  demand   transformers
    $\Lfonly$, we construct a liveness environment \Lv\ for the
    expression $e$.
      %%  If  $e$  is  an expression,  then  $  \Lv  =
    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
    %% application,  then  $  \Lv  =  ref(e,  \sigma,  \Lfonly)$.  
 We  similarly  define  the  liveness  environment  for  the
 expressions $e$ in each of the suspended evaluation contexts $(\rho, \ell,  e, \sigma)$ in $S$, giving
 a stack of liveness environments that we shall denote $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current  context and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff there  is no
    variable  $x$  in either  the  current  environment or  a  stacked
    environment,  say  $\rho$,  such   that  $x.\alpha$  is  live  and
    corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
    domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
    \oxrb = \ell$  and $x.\alpha \in \Lv$, where \Lv\  is the liveness
    environment corresponding to $\rho$ mentioned in step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
the transition is the same as the earlier {\sc car-clo}
rule.

>>>>

\begin{abstract} 
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs. Our approach is to analyze
programs  for liveness  of  heap-allocated data.   The  result of  the
analysis is  used to preserve  only live data---a subset  of reachable
data---during garbage  collection.  The result  is an increase  in the
garbage reclaimed and  a  reduction in  the peak memory requirement 
of programs.   While this  technique has already  been shown  to yield
benefits for  eager first-order  languages, the  lack of  a statically
determinable execution  order and  the presence  of closures  pose new
challenges  for lazy  languages.  These  require changes  both in  the
liveness analysis itself and in the design of the garbage collector.

@@

While  the runtime  system  of most  functional  languages includes  a
garbage  collector    to  efficiently  reclaim  memory,  empirical
studies on Scheme~\cite{karkare06effectiveness} and, more importantly,
on  Haskell~\cite{rojemo96lag}  programs  have shown  that  garbage collectors  leave
uncollected a  large number of  memory objects that are  reachable but
not live (here {\em live} means  the object can potentially be used by
the program  at a  later stage).  This  results in  unnecessary memory
retention.

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis  has  two parts.   We  first  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function  calls  during analysis.   Next,  the  concrete demand  on  a
function body is obtained through a conservative approximation similar
to  0-CFA~\cite{Shivers:1988} that  combines  the demands  on all  the
calls to the function.  The result of the analysis is an annotation of
certain program points with  deterministic finite-state automata (DFA)
capturing the liveness of variables at these points.  Depending on the
program  point where  garbage collection  has been  triggered, the
garbage collector
consults a  set of automata  to restrict reachability  during marking.
This results in an increase  in the garbage reclaimed and consequently
in fewer garbage collections.

While this idea has been shown  to be effective for a first-order {\em
  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
the technique is not possible for lazy languages, where heap-allocated
objects may  include closures (runtime representations  of unevaluated
expressions).   Firstly, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  Moreover,  for liveness-based
garbage collection to  be effective, we need to extend  it to closures
apart from  evaluated data.  Since a  closure can escape the  scope in
which it was created, it has  to carry the liveness information of its
free variables. Moreover, as  execution progresses and possible future
uses are eliminated,  we update the liveness information  in a closure
with a  more precise version.  Because of these reasons,   the garbage
collector  also   becomes  significantly   more  complicated   than  a
liveness-based collector for a eager language.

\warning{Experiments  with  a   single  generation  copying  collector
  (Section~\ref{sec:experiments})  confirm  the  expected  performance
  benefits.   Liveness-based  collection  results in  an  increase  in
  garbage reclaimed.   As a  consequence, there is  a reduction  in the
  number of collections up to ???\% and a decrease in the
  minimum memory requirement up to ???\%. 
  As an added benefit, there is also a reduction in the overall
  execution time for some programs.}

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\input{fig_motiv}
Figure~\ref{fig:mot-example} shows an example program and the state of
the heap at  the program point $\pi$, i.e. just  before the evaluation
of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
node either  represents atomic  values ($\NIL$,  integers etc.),  or a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of  variables  or fields.  The figure shows
the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
triggered by the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.

@@

%\comment{  
Consider the  program in Figure~\ref{fig:mot-example2}.   As mentioned
earlier, a lazy evaluation of the $\LET$ expression at $\pi_1$ creates
a closure  for $(\length~\px)$  instead of  evaluating it.   Since the
closure may escape the scope in which it is created, it carries a copy
of $\px$ within itself.  We treat the  copy of $\px$ in the closure as
being separate from the $\px$ introduced  by the $\LET$, and call it a
\emph{closure  variable}.  For  liveness calculations,  such variables
are distinguished  from variables  introduced by  \LET s  and function
arguments  (called \emph{stack  variables}  since they  reside in  the
activation  stack). Notationally,  we distinguish  a closure  variable
from  its corresponding  stack variable  by subscripting  it with  the
program   point  where   the   closure  was   created\footnote{Mutiple
  occurrences  of the  same  variable in  an  application are  further
  distinguished by their positions in the application.}.

@@

  Closure  variables get  their  liveness values  through  a chain  of
  dependences beginning at  a variable at an evaluation  point.  As an
  example, consider the following program:

\input{example-to-illustrate-liveness}

The  variable \pz\   returned  at  $\psi_2$ depends  on  \py\ 
  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
  closure  variable  $\px_{\pi_1}$   through  $(\length~\px_{\pi_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\pi_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $\pz$  is  evaluated   at  $\psi_2$  due  to  the
  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $\px_{\pi_1}$ to be
  a union of the liveness arising  out of these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $\pz$ at $\psi_2$, it can safely consider only the liveness
  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
  collection  just  before  each  $\rightsquigarrow$  transition.  The
  simulated garbage collection starts from  the root-set and inserts a
  special value  $\bot$ at each location  in the heap that  contains a
  reference that is reachable but not live. Any attempt to dereference
  such  locations during  the  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,}
  $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for \sc{car-clo-bang} and
  \sc{car-clo} \label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial rule is  $\sigma_{all}$ and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it the expression after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environment for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\oxlb\rho(x), \alpha
    \oxrb = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
%%     \oxrb = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
the transition is the similar to the earlier {\sc car-clo}
rule.

Consider a trace  of a minefield execution of a  program $p$, possibly
ending in  a \bang\ state. We  can construct a call-tree  based on the
trace in  which each node represents  a function that was  called (but
did not necessarily  return because of a \bang).  Assume  that each of
the nodes of  the tree is also annotated with  the program point where
the corresponding call  was invoked.  This tree can be  used to inline
function calls in a hierarchical fashion.  The details of the inlining
can be found in~\cite{asati14lgc}.
For  a
call-less program,  the initial state of  the minefield
semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
e_{\mainpgm}$.
